BVP Paper

### 2017-12-08 tex/bvp_paper.tex
restructured paper

Knowledge of Mixing Length theory and the nature of evolved convection has been used in some 
previous studies (e.g., \cite{brandenburg&all2005}) to choose smarter initial conditions than
the hydrostatic polytropic states.  However, these assumptions work best in convective regions
which are bounded by stable layers.  In simple atmospheres where boundary layers form to meet
boundary conditions at the upper and lower edge of the atmosphere, we cannot know 
\emph{a priori} what the extent or shape of the boundary layer will be, as that must be chosen
by the convective dynamics, and specification of the proper boundary layer is essential for
placing the atmosphere along the proper adiabat.

Here we present a method for using simple boundary value problems, along with information about the evolved flow fields,
to fast-forward the slow thermal evolution of convecting simulations.  We present this method in the context of RBC, and
then demonstrate that it applies to stratified, compressible convection under a simple modification.  These methods allow
us to study the convective flows driven by the evolved thermal profile while only requiring initial value problems to run
for long enough to resolve the fast dynamical timescales of convection.

\section{Experiment} 
\label{sec:experiment}
\subsection{Incompressible, Boussinesq \RB Convection}
In studies of convection, it is natural to nondimensionalize the flows on the freefall velocity.
Under the Boussinesq approximation, where $\rho = \rho_0 (1 - \alpha T_1)$ on the gravitational term in the momentum
equation and $\rho = \rho_0$ on all other terms, the freefall velocity is $v_{\text{ff}} = \sqrt{\alpha g L^2 (dT/dz)_0}$.
where $L$ is the depth of the domain, $g$ is the gravitational acceleration, and $(dT/dz)_0$ is the initial temperature
gradient of the atmosphere, which controls the flux through the system.  The typical timescale and length scale
then relate to each other according to $t_{\text{ff}} = L/v_{\text{ff}}$.
Under this nondimensionalization, the Boussinsq, incompressible equations of motion are
\begin{gather}
\DivU = 0, 
	\label{eqn:incompressible}
\\
\frac{\partial \bm{u}}{\partial t} + \bm{u}\cdot\grad\bm{u} =
-\grad\varpi + T_1\hat{z} + \frac{\text{Pr}}{\text{Ra}}\grad^2\bm{u}, 
	\label{eqn:bouss_momentum_eqn}
\\
\frac{\partial T_1}{\partial t} + \bm{u}\cdot\grad(T_0 + T_1) = \frac{1}{\text{Pr}\,\text{Ra}}\grad^2 T_1,
	\label{eqn:bouss_energy_eqn}
\end{gather}
where
\begin{equation}
\text{Ra} = \frac{g \alpha L^4 \left(\frac{dT}{dz}\right)_0}{\nu\chi} = \frac{(L\,v_{\text{ff}})^2}{\nu\chi}, \qquad \text{Pr} = \frac{\nu}{\chi}
\end{equation}
are the nondimensional Rayleigh and Prandtl numbers, where $\nu$ is the viscous diffusivity and $\chi$ is
the thermal diffusivity.

One hallmark of Boussinesq convection is that upflows and downflows are symmetric, and thus the long time- and horizontal-
average of the velocity field is exactly zero.  Assuming that the velocity field has perfect symmetry over a long time
average, the horizontally-averaged, steady state momentum equation is just
\begin{equation}
\frac{\partial}{\partial z}\angles{\varpi} = \angles{T_1}\hat{z},
	\label{eqn:simple_momentum}
\end{equation}
where $\angles{A}$ represent the horizontal- and time- average of the quantity $A$.  Thus, in the evolved solution, the
vertical profile of the pressure perturbations are directly determined by the temperature field. 
Under the Boussinesq approximation, $\bm{u}\cdot\grad(A) = \Div{\bm{u}A}$, and thus the steady state energy equation
is just
\begin{equation}
\frac{\partial}{\partial z}\angles{w(T_0 + T_1)} = \frac{1}{\text{Pr}\text{Ra}}\frac{\partial^2}{\partial z^2} \angles{T_1}.
	\label{eqn:simple_energy}
\end{equation}
Eqns (\ref{eqn:simple_momentum}) \& (\ref{eqn:simple_energy}) combine to form a second-order ODE which specifies the
mean vertical profile of the temperature and pressure fields in the evolved solution.  These fields are completely specified
by a set of two thermal boundary conditions and the evolved 
profile of the convective flux, $F_{\text{conv}} = \angles{w(T_0 + T_1)}$.

Under the choice of mixed thermal boundary conditions ($dT_1/dz = 0$ at $z = -1/2$, $T_1 = 0$ at $z = +1/2$), the initial
atmosphere starts off with much more thermal energy ($\propto T$) than the evolved adiabatic solution pegged by the
top temperature boundary condition.  This excess energy must leave the system through the upper (fixed $T$) boundary
during the convective transient, and this results in an asymmetric flux profile during the slow thermal evolution of the
atmosphere.  Furthermore, the convective flux (and the flux at the upper boundary layer) are O(1) during the convective
transient, whereas the flux entering the atmosphere at the lower boundary is O(Ra$^{-1/2}$), so the asymmetry of the fluxes
becomes increasingly pronounced as Ra is increased into the turbulent realm.

In order to find the evolved temperature profile of the atmosphere, the evolved profile of the convective flux must
be known.  We know that the evolved solution should be in flux equilibrium, and we know that the evolved flux through
the atmosphere is the flux entering through the bottom.  Thus, the steady-state profile of the convective flux can be
approximated as
\begin{equation}
F_{\text{conv, steady}} = F_{\text{bot}}\frac{\angles{w(T_0 + T_1)}}{\angles{w(T_0 + T_1) - \kappa \partial_z (T_0 + T_1)}}
= F_{\text{bot}}\frac{\angles{F_{\text{conv, IVP}}}}{\angles{F_{\text{tot, IVP}}}}.
\end{equation}
Or, put simply, the steady state convective flux is what you get get rid of the asymmetry in the flux profile.
This flux defined here is used in a 1D boundary value problem solve of 
eqns (\ref{eqn:simple_momentum}) \& (\ref{eqn:simple_energy}) along with the appropriate boundary values.  The solution
to that BVP is then used as new initial conditions, convection restarts, and so on.

\subsection{Fully Compressible Convection}
We study stratified convection in an ideal gas whose adiabatic index is $\gamma = 5/3$.
The initial atmospheric stratification is polytropic \cite{anders&brown2017}.
We assume a Newtonian radiative conduction term \cite{lecoanet&all2014}, and solve the
fully compressible Navier-Stokes equations of the form
\begin{gather}
\frac{\partial \ln \rho}{\partial t} + \bm{u}\cdot\grad\ln\rho + \DivU = 0
	\label{eqn:FC_continuity}
\\
\frac{D \bm{u}}{Dt} = -T\grad\ln\rho - \grad T + \bm{g} - \Div{\stressT}
	\label{eqn:FC_momentum}
\\
\frac{D T}{D t} + (\gamma-1)T\DivU = 
\frac{1}{\rho c_V}\left(\kappa\grad^2 T - [\stressT\cdot\grad]\cdot\bm{u}\right),
	\label{eqn:FC_energy}
\end{gather}
where $D/Dt \equiv \partial/\partial t + \bm{u}\cdot\grad$ and the viscous stress
tensor is defined as
\begin{equation}
\Pi_{ij} \equiv -\mu\left(\frac{\partial u_i}{\partial x_j} + \frac{\partial u_j}{\partial x_i} - \frac{2}{3}\delta_{ij
}\DivU\right)
\label{eqn:stress_tensor}
\end{equation}
and $\delta_{ij}$ is the Kronecker delta function.

\subsubsection{The Boundary Value Equations}
In studies of fully compressible convection, the flux carried by the adiabatic temperature gradient
is not available for convection.  Thus, only the flux \emph{in excess} of the adiabat will drive
convection and be carried by convection.  As such, this is the only portion of the flux which
must be examined to determine if the solution is in a converged state.  In a system with
a fixed flux boundary condition, the available superadiabatic flux is
\begin{equation}
F_{\text{avail}} = -\kappa(\grad T_0 - \grad T_{\text{ad}}) = \kappa\frac{\epsilon}{c_P}\grad T_0,
\end{equation}
which is small when $\epsilon$ is small and only requires low Mach number convective flows
to carry it.  In a perfectly evolved solution, there will be thin boundary layers in which
conduction carries this flux in addition to the adiabatic flux, but in an efficient convective
interior, convective fluxes must carry this full amount.

\section{Results \& Discussion}
\label{sec:results}
Here we talk about how the solutions are different, or similar.  This includes:
\begin{enumerate}
\item Showing that the flow fields look similar
\item Showing how the temperature / flux profiles look similar/different
\item showing how Nu and Re scale with Ra in BVP / IVP.
\item showing how the PDFs of $w$, $wT$, and $T$ change.
\end{enumerate}

@@

>>>>

Knowledge of Mixing Length theory and the nature of evolved convection has been used in some 
previous studies (e.g., \cite{brandenburg&all2005}) to choose initial conditions which are
closer to the correct evolved adiabat than
the hydrostatic polytropic states.  However, these assumptions work best in convective regions
which are bounded by stable layers.  In simple atmospheres where boundary layers form to meet
boundary conditions at the upper and lower edge of the atmosphere, we cannot know 
\emph{a priori} what the extent or shape of the boundary layer will be, as that must be chosen
by the convective dynamics, and specification of the proper boundary layer is essential for
placing the atmosphere along the proper adiabat.

Here we present a method for using simple boundary value problems, 
along with information about the evolved flow fields,
to fast-forward the slow thermal evolution of convecting simulations.  
We present this method in the context of RBC, and
then demonstrate that applying it to stratified, compressible convection 
is simple.  These methods allow us to study the convective flows driven by 
nearly the evolved thermal profile while only requiring initial value problems to run
for long enough to resolve the fast dynamical timescales of convection.

\section{Incompressible, Boussinesq \RB Convection}
\label{sec:experiment}
In general, the method for fast-forwarding the atmospheric structure to the correct adiabat is
simple and involves only a few steps.  At the start of convective simulations, a large transient
occurs while the hydrostatic state gives way to the convective state.  After the peak of this
transient, we begin to average the mean vertical profile of the convective flux and the conductive
flux through the domain.  Once those flux profiles are converged, we solve a simple 
boundary value problem using the information that those fluxes provide to determine what the
evolved thermal structure of the atmosphere can be.  The stratification of the initial value
problems is then adjusted to match the output of the BVP, and we run convection for dynamical
timescales to let the flows adjust to the new, updated thermal structure.  We then measure
quantities of the dynamics, and compare them to the dynamics of simulations which underwent a
long thermal rundown, to determine if our simple BVPs effectively get our convective solution
to the correct state.

\subsection{Governing Equations, Nondimensionalization, and Domain setup}
In our first set of experiments, we adopt the Oberbeck-Boussinesq approximation.  Here, the
fluid has constant kinematic viscosity ($\nu$), thermal diffusivity ($\kappa$), and coefficient
of thermal expansion ($\alpha$).  We non-dimensionalize length by the layer height ($L_z$) ,
temperature by the (constant) initial temperature gradient across the layer ($\grad T_0$), and time
by the freefall timescale ($L_z / v_{\text{ff}}$, with $v_{\text{ff}} = \sqrt{\alpha g L_z^2 \grad T_0}$, where $g$ is 
uniform gravitational acceleration in the $-\hat{z}$ direction). The dimensionless Boussinesq
equations governing the velocity $\bm{u} = u\hat{x} + v\hat{y} + w\hat{z}$, temperature
$T = T_0 + T_1$, and reduced pressure $\varpi$ are (cite Rayleigh, chandrasekar)
\begin{gather}
\DivU = 0, 
	\label{eqn:incompressible}
\\
\frac{\partial \bm{u}}{\partial t} + \bm{u}\cdot\grad\bm{u} =
-\grad\varpi + T_1\hat{z} + \frac{\text{Pr}}{\text{Ra}}\grad^2\bm{u}, 
	\label{eqn:bouss_momentum_eqn}
\\
\frac{\partial T_1}{\partial t} + \bm{u}\cdot\grad(T_0 + T_1) = \frac{1}{\text{Pr}\,\text{Ra}}\grad^2 T_1,
	\label{eqn:bouss_energy_eqn}
\end{gather}
where the dimensionless control parameters are the Rayleigh and Prandtl numbers,
\begin{equation}
\text{Ra} = \frac{g \alpha L^4 \left(\frac{dT}{dz}\right)_0}{\nu\chi} = \frac{(L\,v_{\text{ff}})^2}{\nu\chi}, \qquad \text{Pr} = \frac{\nu}{\chi}.
\end{equation}
The dimensionless vertical extent of the domain is $z = [-1/2, 1/2]$, and at the boundaries
we impose no-slip, impenetrable boundary conditions such that $w = u = v = 0$ at $z = \pm 1/2$.
At the lower boundary, we employ a fixed flux condition such that $\partial T_1 / \partial z = 0$
at $z = -1/2$, and we impose a fixed temperature condition $T_1 = 0$ at $z = 1/2$. Both
horizontal directions are periodic and have equal aspect ratio, $\gamma = 4$, such that
the horizontal coordinates $x, y = [0, \Gamma]$.

\subsection{The Boundary Value Problem}
The Boussinesq BVP is simply the time-stationary, horizontally- and time- averaged equations
of hydrostatic balance and energy conservation,
\begin{gather}
\frac{\partial}{\partial z}\angles{\varpi} = \angles{T_1}\hat{z},
	\label{eqn:bouss_BVP_momentum}
\\
\frac{\partial}{\partial z}\angles{wT_1} = \frac{1}{\text{Pr}\text{Ra}}\frac{\partial^2}{\partial z^2} \angles{T_1},
	\label{eqn:bouss_BVP_energy}
\end{gather}
where $\angles{A}$ represents a time- and horizontally averaged profile of the quantity $A$.  These
equations arise from taking time- and horizontal- averages of Eqns (\ref{eqn:bouss_momentum_eqn}-\ref{eqn:bouss_energy_eqn})
and neglecting terms that vanish due to symmetry in the evolved flows.  Convective flows
are perturbations around a thermal profile defined by these equations in the proper evolved state.

In Boussinesq RBC, the thermal structure of the atmosphere is fully determined by the specification
of the convective flux, $F_{conv} = \angles{w T_1}$.  If this profile is known, then $T_1$ and
$\varpi$ can be found under the proper specifications of boundary conditions.  
Under the choice of mixed thermal boundary conditions, the initial
atmosphere contains more thermal energy ($\propto T$) than the 
evolved adiabatic solution.  As the atmosphere adjusts to be nearly isothermal in the interior,
it must evolve towards the (cold) temperature value specified at the upper boundary.
The evolution of the atmosphere results in an asymmetric flux profile during 
the slow thermal evolution of the atmosphere.  Furthermore, under our nondimensionalization,
the convective flux (and the flux at the upper boundary layer) are O(1) during the convective
transient, whereas the flux entering the atmosphere at the lower boundary is O(Ra$^{-1/2}$), so the asymmetry of the fluxes
becomes increasingly pronounced as Ra is increased into the turbulent realm.

In order to find the evolved temperature profile of the atmosphere using the Boussinesq BVP equations, 
the evolved profile of the convective flux must be properly specified.  In order to construct
this profile, we acknowledge that the evolved solution will in flux equilibrium, 
carrying the amount of the flux entering through the bottom.  
Thus, the steady-state profile of the convective flux can be approximated as
\begin{equation}
F_{\text{conv, steady}} = F_{\text{bot}}\frac{\angles{wT_1}}{\angles{wT_1 - \kappa \partial_z (T_0 + T_1)}}
= F_{\text{bot}}\frac{\angles{F_{\text{conv, IVP}}}}{\angles{F_{\text{tot, IVP}}}}.
\label{eqn:bouss_BVP_fconv}
\end{equation}
Or, put simply, the steady state convective flux is retrieved by properly removing the asymmettry
from the flux profile.

In our Boussinesq BVPs, we solve Eqns. (\ref{eqn:bouss_BVP_momentum}-\ref{eqn:bouss_BVP_energy}),
substituting $\angles{wT_1} = F_{\text{conv, steady}}$ as defined in eqn. (\ref{eqn:bouss_BVP_fconv})
to retrieve the proper vertical profile of $T_1$ and $\varpi$.  We then update the mean horizontal value
of $T_1$ and $\varpi$ in a corresponding IVP, and continue to timestep forward with the newly adjusted
atmosphere.

\subsection{Numerics}

\subsection{Results}
Here we talk about how the solutions are different, or similar.  This includes:
\begin{enumerate}
\item Showing that the flow fields look similar
\item Showing how the temperature / flux profiles look similar/different
\item showing how Nu and Re scale with Ra in BVP / IVP.
\item showing how the PDFs of $w$, $wT$, and $T$ change.
\end{enumerate}

@@

\section{Fully Compressible Convection}

\subsection{Governing Equations}
We study stratified convection in an ideal gas whose adiabatic index is $\gamma = 5/3$.
The initial atmospheric stratification is polytropic \cite{anders&brown2017}.
We assume a Newtonian radiative conduction term \cite{lecoanet&all2014}, and solve the
fully compressible Navier-Stokes equations of the form
\begin{gather}
\frac{\partial \ln \rho}{\partial t} + \bm{u}\cdot\grad\ln\rho + \DivU = 0
	\label{eqn:FC_continuity}
\\
\frac{D \bm{u}}{Dt} = -T\grad\ln\rho - \grad T + \bm{g} - \Div{\stressT}
	\label{eqn:FC_momentum}
\\
\frac{D T}{D t} + (\gamma-1)T\DivU = 
\frac{1}{\rho c_V}\left(\kappa\grad^2 T - [\stressT\cdot\grad]\cdot\bm{u}\right),
	\label{eqn:FC_energy}
\end{gather}
where $D/Dt \equiv \partial/\partial t + \bm{u}\cdot\grad$ and the viscous stress
tensor is defined as
\begin{equation}
\Pi_{ij} \equiv -\mu\left(\frac{\partial u_i}{\partial x_j} + \frac{\partial u_j}{\partial x_i} - \frac{2}{3}\delta_{ij
}\DivU\right)
\label{eqn:stress_tensor}
\end{equation}
and $\delta_{ij}$ is the Kronecker delta function.

\subsection{The Boundary Value Equations}
In studies of fully compressible convection, the flux carried by the adiabatic temperature gradient
is not available for convection.  Thus, only the flux \emph{in excess} of the adiabat will drive
convection and be carried by convection.  As such, this is the only portion of the flux which
must be examined to determine if the solution is in a converged state.  In a system with
a fixed flux boundary condition, the available superadiabatic flux is
\begin{equation}
F_{\text{avail}} = -\kappa(\grad T_0 - \grad T_{\text{ad}}) = \kappa\frac{\epsilon}{c_P}\grad T_0,
\end{equation}
which is small when $\epsilon$ is small and only requires low Mach number convective flows
to carry it.  In a perfectly evolved solution, there will be thin boundary layers in which
conduction carries this flux in addition to the adiabatic flux, but in an efficient convective
interior, convective fluxes must carry this full amount.

\subsection{Results}

\section{Discussion \& Conclusions}
\label{sec:results}
